In-silico molecular enrichment and clearance of the human intracranial space

### 2024-12-19 paper/supplementary_main.tex
Fix in A.1.4.

In addition, we consider two sets of perivascular networks: a
periarterial network $\Lambda_a$ represented by the (connected)
centerlines $\Lambda^i_{(a)}$ of the arterial tree, and a perivenous
network $\Lambda_v$ associated with the centerlines $\Lambda^i_{(v)}$
of the veins. Recall that we represent the vascular domains as the
union of cylindrical vessels of radius $R_1^i$ surrounding the
centerlines $\Lambda^i$. Moreover, we consider the surrounding
perivascular spaces as the union of annular cylinders $\Omega^i$ of
inner radius $R_1^i$ and outer radius $R_2^i > R_1^i$ and thus of
width $R_2^i - R_1^i$. \discuss{We denote the outer lateral surface of the
periarterial and perivenous spaces by $\Gamma_a$ and $\Gamma_v$,
respectively.} We omit the subscript $a$ or $v$ when referring to any
such network, vessel segment or perivascular outer surface. We assume
that $\Lambda$ is parametrized by the coordinate $s$, and with a minor
abuse of notation, simply write $s \in \Lambda$ to represent the point
${\bm \lambda}(s)$ on $\Lambda$ corresponding to $s$. Finally, we view
each perivascular network both as a geometric domain and as a directed
graph with the centerlines $\{ \Lambda^i \}$ as oriented edges and the
connections as nodes $\mathcal{V}$.

@@

%\mk{What is justification of \eqref{eq:1d_bcs}?}
%% Both $\Omega$ and $\Lambda$ are fixed in time and we assume
%% constant-in- time radii (despite varying radii to drive peristaltic
%% flow, see \cref{sec:sup:peristalsis}).

@@

\subsubsection{Steady flow in perivascular networks -- induced by pressure differences}

To model steady flow induced by pressure differences between end nodes
in a perivascular network $\Lambda$, with edges $\{\Lambda^i\}_i$,
internal nodes $\mathcal{B}$ and end nodes $\mathcal{N}$, we consider
the following system of hydraulic network
equations~\cite{tithof2022network, daversin2022geometrically,
  gjerde2024directional}. The unknowns are the PVS flux $\hat{q} :
\Lambda \rightarrow \R$ and pressure $\hat{p} : \Lambda
\rightarrow \R$, which represent the fluid flux and the average
pressure across cross-sections of the PVSs, respectively. These are
defined for each $\Lambda^i$ by solving:
\begin{subequations}
\begin{alignat}{2}
  \hat{q} + \frac{A \kappa}{\mu} \partial_{s} \hat{p} & = 0 &&
  \quad \text{ in } \Lambda^i , \\ 
  - \partial_s \hat{q} & = 0 &&
  \quad \text{ in } \Lambda^i ,  
\end{alignat} 
\label{eq:1d_flow}%
\end{subequations}%
where $A$ is the area of the PVS cross-sections, $\mu$ is the CSF viscosity as before, and $\kappa$ is derived from an assumption of Poiseuille flow in the annular cross-section of the PVS as~\cite{tithof2022network}:  \rami{Here,  we consider the steady state  case and we assume that the PVS domain is fixed and not pulsating  }
\begin{equation}
\kappa = \frac18 \left( R_2^2 + R_1^2 - \frac{1}{\ln(R_2/R_1)} (R_2^2- R_1^2) \right). 
\end{equation}
The PVS flux $\hat{q}$ and the associated average PVS velocity in the
axial direction $\hat{u}$ are directly related by the cross-sectional area
\begin{equation}
  \hat{q}^i = A^i \hat{u}^i \qquad \foralls \Lambda^i.
\end{equation}

In addition, to complete~\eqref{eq:1d_flow}, we impose continuity of
the fluid pressure and conservation of flux at each internal node. Write
$\hat{p}^i = \hat{p}|_{\Lambda^i}$. These conditions then read as: for
each $y \in \mathcal{B}$ with connected edges $\mathcal{E}(y)$:
\begin{subequations}
\begin{align}
  \hat{p}^i(y) &= \hat{p}^j(y) \qquad \foralls \Lambda^i, \Lambda^j \in \mathcal{E}(y), \\
  \sum_{\Lambda^i \in \mathcal{E}(y)} \hat{q}^i n^i(y) &= 0  .
\end{align}
\label{eq:1d_flow:bifurcations}
\end{subequations}
where $n^i$ is as defined by~\eqref{eq:def:sign}. Finally, to drive this flow, we impose given fluid pressures $p_0$ at the
network end nodes:
\begin{equation}
  \hat{p}(y) = p_0(y) \qquad y \in \mathcal{N} .
  \label{eq:1d_flow:bcs}
\end{equation}
In the case where modelling flow in the perivascular networks
$\Lambda_a, \Lambda_v$ induced by CSF production, we impose the fluid
pressure $p_{\rm CSF}$ computed in the CSF spaces $\Omega_{\rm CSF}$
(and its harmonic extensions into $\Omega_{\rm PAR}$) as the given
fluid pressures $p_0$.

@@

\subsection{Estimating net perivascular flow induced by peristaltic waves}
\label{sec:sup:peristalsis}

@@

\input{tikz/fig_pvs_flow}
This theoretical formalism~\cite{gjerde2023directional} is defined
relative to a network in the form of a directed, bifurcating tree with
a single supply node/root $i_0$. To extend to a network of cerebral
arteries with multiple supply nodes (such as the basilar and two
internal carotid arteries in the current data
set~\cite{hodneland2019new}), we separate the network $\Lambda$ into
edge-disjoint subnetworks $\Lambda^1, \Lambda^2, \Lambda^3$, one for
each of the supply nodes (\Cref{fig:sup:peristalsis}). Each node is assigned to the subnetwork associated with the nearest supply node, and edges between nodes are preserved. Next, we compute a minimal, bifurcating and directed tree representation of each subnetwork: $\mathcal{T}^1, \mathcal{T}^2, \mathcal{T}^3$. Each tree $\mathcal{T}^j = \cup E_n^j$ consists of the subset of the nodes from $\Lambda^j$ that have degree 1 (are leaf or root nodes) or degree $3$ (are true bifurcation points), and each path between nodes with degree $\not = 2$ in $\Lambda^j$ is represented in $\mathcal{T}^j$ by an edge $E^j$ with edge length $L$ corresponding to the total length of the original path and edge radius $r_o$ as the average of the path radii. For the few (2) nodes in $\Lambda$ with degree $>3$; i.e.~junctions where a branch splits into more than two downstream branches, only the two longest downstream paths are included in $\mathcal{T}^j$.

>>>>

In addition, we consider two sets of perivascular networks: a
periarterial network $\Lambda_a$ represented by the (connected)
centerlines $\Lambda^i_{(a)}$ of the arterial tree, and a perivenous
network $\Lambda_v$ associated with the centerlines $\Lambda^i_{(v)}$
of the veins. Recall that we represent the vascular domains as the
union of cylindrical vessels of radius $R_1^i$ surrounding the
centerlines $\Lambda^i$. Moreover, we consider the surrounding
perivascular spaces as the union of annular cylinders $\Omega^i$ of
inner radius $R_1^i$ and outer radius $R_2^i > R_1^i$ and thus of
width $R_2^i - R_1^i$. We denote the outer lateral surface of the
periarterial and perivenous spaces by $\Gamma_a$ and $\Gamma_v$,
respectively. We omit the subscript $a$ or $v$ when referring to any
such network, vessel segment or perivascular outer surface. We assume
that $\Lambda$ is parametrized by the coordinate $s$, and with a minor
abuse of notation, simply write $s \in \Lambda$ to represent the point
${\bm \lambda}(s)$ on $\Lambda$ corresponding to $s$. Finally, we view
each perivascular network both as a geometric domain and as a directed
graph with the centerlines $\{ \Lambda^i \}$ as oriented edges and the
connections as nodes $\mathcal{V}$.

@@

@@

\subsubsection{Steady flow in perivascular networks induced by pressure differences}

To model steady flow induced by pressure differences between end nodes
in a perivascular network $\Lambda$, with edges $\{\Lambda^i\}_i$,
internal nodes $\mathcal{B}$ and end nodes $\mathcal{N}$, we consider
the following system of hydraulic network
equations~\cite{tithof2022network, daversin2022geometrically,
  gjerde2024directional}. The unknowns are the PVS flux $\hat{q} :
\Lambda \rightarrow \R$ and pressure $\hat{p} : \Lambda
\rightarrow \R$, which represent the fluid flux and the average
pressure across cross-sections of the PVSs, respectively. These are
defined for each $\Lambda^i$ by solving:
\begin{subequations}
\begin{alignat}{2}
  \hat{q} + \frac{A \kappa}{\mu} \partial_{s} \hat{p} & = 0 &&
  \quad \text{ in } \Lambda^i , \\ 
  - \partial_s \hat{q} & = 0 &&
  \quad \text{ in } \Lambda^i ,  
\end{alignat} 
\label{eq:1d_flow}%
\end{subequations}%
where $A$ is the area of the PVS cross-sections, $\mu$ is the CSF viscosity as before, and $\kappa$ is derived from an assumption of Poiseuille flow in the annular cross-section of the PVS as~\cite{tithof2022network}: 
\begin{equation}
\kappa = \frac18 \left( R_2^2 + R_1^2 - \frac{1}{\ln(R_2/R_1)} (R_2^2- R_1^2) \right). 
\end{equation}
The PVS flux $\hat{q}$ and the associated average PVS velocity in the
axial direction $\hat{u}$ are directly related by the cross-sectional area
\begin{equation}
  \hat{q}^i = A^i \hat{u}^i \qquad \foralls \Lambda^i.
\end{equation}

In addition, to complete~\eqref{eq:1d_flow}, we impose continuity of
the fluid pressure and conservation of flux at each internal node. Write
$\hat{p}^i = \hat{p}|_{\Lambda^i}$. These conditions then read as: for
each $y \in \mathcal{B}$ with connected edges $\mathcal{E}(y)$:
\begin{subequations}
\begin{align}
  \hat{p}^i(y) &= \hat{p}^j(y) \qquad \foralls \Lambda^i, \Lambda^j \in \mathcal{E}(y), \\
  \sum_{\Lambda^i \in \mathcal{E}(y)} \hat{q}^i n^i(y) &= 0  .
\end{align}
\label{eq:1d_flow:bifurcations}
\end{subequations}
where $n^i$ is as defined by~\eqref{eq:def:sign}. Finally, to drive this flow, we impose given fluid pressures $p_0$ at the
network end nodes:
\begin{equation}
  \hat{p}(y) = p_0(y) \qquad y \in \mathcal{N} .
  \label{eq:1d_flow:bcs}
\end{equation}
In the case where modelling flow in the perivascular networks
$\Lambda_a, \Lambda_v$ induced by CSF production, we impose the fluid
pressure $p_{\rm CSF}$ computed in the CSF spaces $\Omega_{\rm CSF}$
(and its harmonic extensions into $\Omega_{\rm PAR}$) as the given
fluid pressures $p_0$. Note that we here consider steady PVS flow in a
non-moving domain; the effect of domain motion is addressed
in~\Cref{sec:app:peristalsis}.

@@

\subsection{Estimating net perivascular flow induced by peristaltic waves}
\label{sec:app:peristalsis}

@@

\input{tikz/fig_pvs_flow}
This theoretical formalism~\cite{gjerde2023directional} is defined
relative to a network in the form of a directed, bifurcating tree with
a single supply node/root $i_0$. To extend to a network of cerebral
arteries with multiple supply nodes (such as the basilar and two
internal carotid arteries in the current data
set~\cite{hodneland2019new}), we separate the network $\Lambda$ into
edge-disjoint subnetworks $\Lambda^1, \Lambda^2, \Lambda^3$, one for
each of the supply nodes (\Cref{fig:app:peristalsis}). Each node is assigned to the subnetwork associated with the nearest supply node, and edges between nodes are preserved. Next, we compute a minimal, bifurcating and directed tree representation of each subnetwork: $\mathcal{T}^1, \mathcal{T}^2, \mathcal{T}^3$. Each tree $\mathcal{T}^j = \cup E_n^j$ consists of the subset of the nodes from $\Lambda^j$ that have degree 1 (are leaf or root nodes) or degree $3$ (are true bifurcation points), and each path between nodes with degree $\not = 2$ in $\Lambda^j$ is represented in $\mathcal{T}^j$ by an edge $E^j$ with edge length $L$ corresponding to the total length of the original path and edge radius $r_o$ as the average of the path radii. For the few (2) nodes in $\Lambda$ with degree $>3$; i.e.~junctions where a branch splits into more than two downstream branches, only the two longest downstream paths are included in $\mathcal{T}^j$.
